\chapter{Present and future ground-based detectors}

\author[Giacomo Ciani and Paul Fulda]{G.Ciani and P.Fulda\footnote{To be published in \textit{An overview of Gravitational Wave Theory and Detection.}, eds. G. Auger and E. Plagnol (World Scientific, 2016).}}
\address{University of Florida, Gainesville, FL 32611, USA}


\bigtoc

\tableofcontents

\body


\newpage
\begin{abstract}
	\textit{\textbf{Abstract}:
%		the direct detection of gravitational waves by the Advanced LIGO project on the centenary of their prediction by Einstein was a milestone event  for general relativity, gravitational physics, astrophysics and many other related fields.
%		Given the astonishingly small amplitude of the gravitational waves emitted by even the strongest sources, building an instrument sensitive enough to detect these tiny ripples in space-time was deemed impossible by Einstein himself.
%		Only the bright ideas of visionary physicists and many decades of intense scientific research and technological development have made the feat possible.
		In this chapter, we first briefly review the early history of gravitational wave detection, and how the research turned towards the large scale interferometers that proved to be the most effective devices for gravitational wave astronomy.
		We provide an overview of what is generically defined the ``first generation'' of interferometric detectors, a number of instruments built around the globe at the end of the last century, and differing in size, technology and scope.
		A more detailed description is dedicated to the ``second generation'', or ``advanced detectors'', and their main subsystems, which represent the state of the art of the science and technology in the filed.
		In the final section, we give a glimpse of what the next generation of detectors may look like.
		Given the many similitudes between different projects, throughout the chapter we will use LIGO as the leading example, and highlight how the other projects compared to it both technologically and strategically. This choice is mainly due to the size of the LIGO project and of the LIGO Scientific Collaboration, and to the pivotal role they played in the history of the field and eventually in the first detection of gravitational waves. It is important to note, however, that the comparatively small amount of space and detail dedicated to other endeavors is not indicative of the importance of the role they have played and still play in the field.
		}
\end{abstract}


\body


\input{history.tex} 

\input{1stgen.tex} 

\input{2ndgen.tex} 

\input{future.tex} 

\bibliographystyle{ws-rv-van}    % Bibliography: Author-Date system
\bibliography{presentandfuture}      % pls. call your database here